Variance reduction for Markov chains via martingale representations

### 2020-04-14 aap/Revision/vr-mcmc-revision-new-new.tex
update vr-mcmc-revision-new-new.tex

Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
where for all $y \in \rset^d$, 
\label{eq:definition-bar-a-0}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right] \,, \text{for $r, k\in \mathbb{N}$.}

@@

for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.

@@

The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. One way to estimate them is to use nonparametric regression.
We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions
$A_{q,k}$ for $q=0,\ldots,n-1,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)

@@

Then we solve  for $k=1,\ldots,K,$  the least squares optimization problems

@@

\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+N}^{(s)})\phi_k(\xi^{(s)}_{N+1}) - \psi(X^{(s)}_{N})\right|^2

@@

where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are i.i.d. random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see \cite[Section~11]{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
\label{eq:trunc}
\widehat{A}_{q,k}(y):=
\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq W_k,
W_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}.
 for a sequence of positive real numbers $W_k,$ $k=1,\ldots,K.$
Upon estimating  the functions \((A_{q,k})\), one can construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form

@@

First we note that from  numerical point of view another equivalent representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
In the homogeneous case $\Phi_l=\Phi$, the coefficients $(\bar{a}_{r,k})$ in \eqref{eq:definition-bar-a-0} are given respectively  for all $l\geq 1,$ by
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}

@@

The coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) are obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\bigl\{\bigl(X_1^{(s)},\ldots,X_{N+n}^{(s)}\bigr)\bigr\}_{s=1}^{T}\). More precisely  we define $\widehat{Q}_0(x) = f(x)$ and

@@

\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{l_0+r}^{(s)}) - \psi(X_{l_0}^{(s)})\right|^2

@@

for \(1 \leq l_0+r \leq n-1,\) where $\Psi_m$ is a class of polynomials

>>>>

\begin{cor}
Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
\begin{eqnarray*}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for all $y \in \rset^d$, 
\begin{equation}
\label{eq:definition-bar-a-0}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right] \,, \text{for $r, k\in \mathbb{N}$.}
\end{equation}
\end{cor}
Another equivalent representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful to construct estimators:
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
In the homogeneous case $\Phi_l=\Phi$, the coefficients $\bar{a}_{r,k}$ in \eqref{eq:definition-bar-a-0} are given respectively  for all $l\geq 1,$ by
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right] \,
\quad \text{with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$}
\end{equation}
\end{prop}

@@

for any  function $g \in L^2(\pi).$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.

@@

The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. 
One way to estimate them is to use nonparametric regression.
%%%%% Old Version
%We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions
%$A_{q,k}$ for $q=0,\ldots,n-1,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
%(the so-called ``training paths''):
%\[
%\TrainSet=\Bigl\{(X^{(s)}_{1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
%\]
%with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
%Then we solve  for $k=1,\ldots,K,$  the least squares optimization problems
%\begin{equation}\label{eq:06042018a1}
%\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+N}^{(s)})\phi_k(\xi^{(s)}_{N+1}) - \psi(X^{(s)}_{N})\right|^2
%\end{equation}
%where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are i.i.d. random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see \cite[Section~11]{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
%\label{eq:trunc}
%\widehat{A}_{q,k}(y):=
%\begin{cases}
%\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq W_k,
%\\
%W_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}.
%\end{cases}
% for a sequence of positive real numbers $W_k,$ $k=1,\ldots,K.$
%Upon estimating  the functions \((A_{q,k})\), one can construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
%\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n} \widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\,.
%%%% Old Version
We present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions
$Q_r$ for $r=0,\ldots,n-1,$, defined in \eqref{eq:definition-bar-a}. 
We first generate \(\NtrainPath\)  paths  of the chain \(X\)

@@

Then we solve  the least squares optimization problems

@@

\hat{Q}_{r} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left|   f(X_{r+N}^{(s)}) - \psi(X^{(s)}_{N})\right|^2

@@

where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\).
We then set 
\label{eq:definition-hat-a-A}
\hat{a}_{r,k}(x)= \int \hat{Q}_r(\Phi(x,z)) \phi_k(z) P_\xi(\rmd z) \quad \text{and}
\quad 
\hat{A}_{q,k}(x) = \sum_{r=0}^q \hat{a}_{r,k}(x) \eqsp,
and construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form

@@

@@

The coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) are obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\bigl\{\bigl(X_1^{(s)},\ldots,X_{N+n}^{(s)}\bigr)\bigr\}_{s=1}^{T}\). More precisely  we define $\widehat{Q}_0(x) = f(x)$ and for \(1 \leq r \leq n-1,\) we set 

@@

\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{N+r}^{(s)}) - \psi(X_{N}^{(s)})\right|^2

@@

 where $\Psi_m$ is a class of polynomials
